---
title: "IPW and AIPW"
output:
  html_document:
    css: custom.css
    includes:
      in_header: header.html
header-includes:
   - \usepackage{amsmath}
---



# IPW and AIPW

```{r, echo = FALSE, include = FALSE}
#################################
#
# Author: VS
# Last Revised: Jan 18, 2024
# Keywords: Heterogeneous Treatment Effects
# using Lasso, GRF
#
#################################

rm(list = ls())

# making table
library("kableExtra")

# Helper packages
library(dplyr) # for data wrangling
library(ggplot2) # for graphics
library(splines)

# Modeling packages 
library(ranger)
library(caret) # for automating the tuning process
library(grf)
library(glmnet) # for implementing regularized regression
```


The target is to estimate the average treatment effect (ATE): 

\begin{equation}
\label{eq:ATE}
ATE = E[Y_i(1) - Y_i(0)]...... \; equation (1)  
\end{equation}

Note that using the following two assumptions: 

i) $W_i \perp \{Y_i(0), \; Y_i(1)\}$ (independence assumption)

ii) $Y_i(W) = Y_i$ (SUTVA)

the ATE estimate $\hat{\tau}$ can be written as the difference-in-means estimator: 

\begin{equation}
\label{eq:ATE_estimator}
\hat{\tau} = \frac{1}{N_T} \sum_{W_i = 1} Y_i - \frac{1}{N_C} \sum_{i \in W_i = 0} Y_i 
\end{equation}

where $N_T$ and $N_C$ are the number of treated and control units, respectively. 


In the previous lecture, we disscussed randomized control trial as an ideal approach to estimate ATE. 
In a randomized controlled trial each unit has an equal probability of receiving the treatment. This means the following: 

\begin{equation}
P(W_i = 1 \; | \;  Y_i(0), \;  Y_i(1),  \;  n_T) = \frac{n_T}{n}, \; \; i = \{1, ...., n\} (\#eq:ptreat)
\end{equation}



In equation \@ref(eq:ptreat), $n_T$ refers to the number of units that should receive the treatment.^[Although it is generally recommended to assign half of the sample to the treatment group and the other half to the control group, this is not a strict requirement.] 
In an easy to understand set-up, if a researcher wants $P(W_i = 1) = 0.5$ (unit is equally likely to be treated or untreated), a coin flip can feasibly be used as a mechanism to assign treatment.^[Of course, this is quicky going to be inefficient as the sample size increases. In general, treatment assignment is determinted by a statistical process via a software. For example, if a researcher wants about one-third of the sample treated then a bernoulli trial with the probability of success of 0.33 can be used.] 


Although randomized controlled trials (RCTs) are often considered the gold standard in causal inference, but they cannot always be used due to ethical, moral, and monetary reasons. Returning to the example we used in the previous chapter, it is not ethical to demarcate who can attend the tutoring session versus who cannot. This is precisely what an RCT setting does.

In real-world scenarios, tutoring sessions are typically voluntary. Students who regularly attend these sessions may have different baseline (pre-treatment) characteristics compared to those who do not attend. These differences can introduce biases that complicate causal inference in observational studies.


To proceed further, we require more knowledge about the treatment assignment. In other words, we need to understand which variables determine who attends the tutoring sessions. This information is crucial for identifying potential confounders and for applying methods that can help estimate causal effects in observational settings. This brings us to the **unconfoundedness** assumption. 


**Unconfoundedness**: The treatment assignment is as good as random once we control for $X$s. $\{W_i \perp \{Y_i(0), \; Y_i(1)\} | X_i \} \; for \; all \; x \in \chi$.   

As with the tutoring example, the independence assumption (discussed in the previous chapter) is highly unlikely to hold in observational settings. Let's consider the following scenarios:

a. Out of the ten states that are yet to expand Medicaid, eight fall in South. 
b. Cigarette taxes are higher in states with higher anti-smoking sentiments. 
c. Infrastructure development, such as construction of roads, schools, hopitals, are demand-driven. 
d. The list goes on ..


However, if we manage to observe all the $X$s (covariates) that influence the treatment, we can invoke unconfoundedness for causal inference—assuming the stars align and and the universe decides to cooperate.

## A simple example 

Say, you are interested in evaluating the effect of tutoring program initiated following the first exam on grades at an introductory level course. For simplicity, the possible grades are **A** and **B**. However, students who received B on their first exam are more likely to attend the tutoring session. In other words, $P(W_i = 1 | Y_{iFE} = A) < P(W_i = 1 | Y_{iFE} = B)$. In this case, the treatment assignment is correlated with the past grade, which can predict the grade on the second exam. Hence, using equation (2) to estimate effects of the tutoring program will result in biased estimate. 

Since we know that the probability of treatment is influenced by the grade on the first exam, we can estimate the conditional average treatment effect (CATE) and average them using weights to form an estimate of ATE. Let's take a look at the data. 

```{r}


grade_mat  <- function(grade_vec){
    grade  <- matrix(0, nrow =2, ncol = 3)
    grade[, 1]  <- c("Treat", "Control")
    grade[ ,2]  <- c(grade_vec[1], grade_vec[2])
    grade[ ,3]  <- c(grade_vec[3], grade_vec[4])
    return(grade)
}

# Y_iFS == A
grade  <- grade_mat(c(5, 9, 2, 4))
colnames(grade)  <- c("1st Exam = A", "A (2nd Exam)", "B (2nd Exam)")

grade  %>% kable()  %>% 
    kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

grade  <- grade_mat(c(15, 1, 5, 4))
colnames(grade)  <- c("1st Exam = B", "A (2nd Exam)", "B (2nd Exam)")

grade  %>% kable()  %>% 
    kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```


$\hat{\tau}_{FE=A} = \frac{5}{7} - {9}{13} = 2.1 \; pp$

$\hat{\tau}_{FE=B} = \frac{15}{20} - {1}{5} = 55 \; pp$

$\hat{\tau}_{AGG} = \frac{20}{45} \hat{\tau}_{FE=A} - \frac{25}{45} \hat{\tau}_{FE=B} = 31.48 \; pp$. 

The first two are CATEs, which is then averaged to form ATE using appropriate weights on the third line. This simple example using the discreate feature space provides intuition that if variables influencing the treatment assignment are observed then ATE estimate can be uncovered by taking weighted average of CATE estimates (these are also group-wise ATE).^[In this case, CATEs are different across the two sub-groups. Sometimes the core interest of analysis can be uncovering the heterogeneous treatment effects, which motivates estimation and inference on CATEs across two or more sub-groups.]

## Continuous features and propensity score 

Previously we discussed the case of a discrete feature. In this case we estimates group-wise ATE and used the weighted average to obtain ATE. When there are many features, this approach is prone to the curse of dimensionality. Moreover, if features are continuous, we won't be able to estimate ATE at each value of $x \in \chi$ due to lack of enough sample size. Instead of estimating group-wise ATE and averaging them, we would want to use a more indirect approach. This is when propensity score comes in. 

The assumption is that we have collected enough features (discrete, continuous, interactions, higher degree polynomials) to back unconfoundedness. This again means that the treatment assignment is as good as random after controlling for $X_i$. More formally, 

$W_i \perp \{Y_i(0), \; Y_i(1)\} | \; X_i \; ............. equation(3)$ 

Intuitively, after accounting for the features that influence who receives treatment, the treatment assignment is as good as random among the sub-group with similar features. But in actuality we are not interested in splitting groups to estimate group-wise treatment effects.   

**Propensity score $e(x)$.** The probability of being treated conditional on features $X$s. 

$e(x) = P(W_i = 1 | X_i = x) \; ............. equation(4)$. 

The key property of the propensity score is that it balances units in the treatment and control groups. If equation (3) holds, we can write the following: 

$W_i \perp \{Y_i(0), \; Y_i(1)\} | \; e(X_i) \; ............. equation(5)$ 

What equation (5) says is that instead of controlling for $X$ one can control for the probability of treatment $(e(X))$ to establish the desired property that the treatment is as good as random. The implication of equation (5) is that if we partition observations in groups with similar propensity score then we can estimate group-wise treatment effects and aggregate to form an estimate for ATE. This can be done using the propensity score stratification. 

## Propensity score stratification

1. Order observations according to their estimated propensity score. 

$\hat{e}(X)_{i1}, \; \hat{e}(X)_{i2}, ... \; \hat{e}(X)_{iN}$

2. Form $J$ strata of equal size and take the simple difference in mean between the treated and control units within each strata. These are $\hat{\tau}_j$ for $j = \{1, \; 2, \; ..., \; N\}$.  

3. Form the ATE, 

$\hat{\tau}_{Strat} = \frac{1}{J} \sum_{j = 1}^{J} \hat{\tau}_j$

Here, $\hat{\tau}_{Strat}$ is consistent for $\tau$, meaning that $\hat{\tau}_{Strat} \rightarrow_p \tau$ given that $\hat{e}(x)$ is consistent for $e(x)$ and the number of strata grows appropriately with $N$. However, one needs to set the number of strata, which can be a bit ad-hoc.  

## IPW and Estimation

A more natural way to exploit the condition of unconfoundedness is to weight observations by their propensity score, which is known as the inverse probability weighting. As before $\hat{e}(x)$ is defined as an estimated propensity score. 

$\hat{\tau}_{IPW} = \frac{1}{N}\sum_{i = 1}^{N} (\frac{Y_i . W_i}{\hat{e}(X_i)} - \frac{Y_i . (1-W_i)}{1 - \hat{e}(X_i)}) \; ....equation(6)$

Intuitively, observations with high propensity score within the treated group are weighted down, while observations with higher propensity score in the control group are weighted more. In this way, propensity score is used to balance the differences in features across the treatment and control groups. Note that the validity of $\hat{\tau}$ still hinges on the unconfoundedness assumption. 

**Limitation of IPW Estimate.** One way to analyze the accuracy of $\hat{\tau}_{IPW}$ is to compare it with the oracle IPW estimate, $\hat{\tau}_{IPW}^{*}$. The oracle estimate is obtained from the known propensity score. Readers are referred to Wager (2018) lecture notes regarding the details of the comparison. Briefly, comparison between $\hat{\tau}_{IPW}^{*}$ and $\hat{\tau}_{AGG}$ suggests that the oracle IPW under-performs $\hat{\tau}_{AGG}$. In other words, the variance of the oracle estimate is larger than that of $\hat{\tau}_{AGG}$. 

Algorithmically, we can form *score* as:

$(\frac{Y_i \times W_i}{\hat{e}(X_i)} - \frac{Y_i \times (1-W_i)}{1 - \hat{e}(X_i)})$

The mean of it results to $\hat{\tau}$ and the standard error of the estimate is simply $\frac{\hat{\sigma}_{score}}{\sqrt{N}}$.

**Estimation**

```{r}
#################################
# Author: VS
# Last Revised: Jan 16, 2024
# Keywords: IPW, AIPW, GRF
#
#
#
#################################


set.seed(194)

# Generate Data 
n <- 2000
p <- 10
X <- matrix(rnorm(n * p), n, p)
X.test <- matrix(0, 101, p)
X.test[, 1] <- seq(-2, 2, length.out = 101)
W <- rbinom(n, 1, 0.4 + 0.2 * (X[, 1] > 0))
prob  <- 0.4 + 0.2 * (X[, 1] > 0)
Y <- 2.56 * W + X[, 2] + pmax(X[, 1], 0) + rnorm(n)
plot(X[, 1], X[, 2], col = as.factor(W))

#paste0("average treatment effect is: ", round(mean(pmax(X[, 1], 0)), 3))


#################################
#################################
#
# Inverse Probability Weighting
#
#################################
#################################

# use the random forest to get the propensity score 
dat  <- data.frame(W, X)
n_features  <- length(setdiff(names(dat), "W"))

# A. ranger (probability tree)
rf1_ranger  <- ranger(
    W ~ ., 
    data = dat, 
    mtry = min(ceiling(sqrt(n_features) + 20), n_features),  
    num.trees = 2000, 
    probability =  TRUE
)
# OOB predictions from ranger
p.ranger  <- rf1_ranger$predictions[, 1]  

# B. probability tree using GRF

# cross-fitting index 
K  <- 10
ind  <- sample(1:length(W), replace = FALSE, size = length(W))
folds  <- cut(1:length(W), breaks = K, labels = FALSE)
index  <- matrix(0, nrow = length(ind) / K, ncol = K)
for(f in 1:K){
    index[, f]  <- ind[which(folds == f)]
}

# Build RF using GRF P(W = 1 | X)
fun.rf.grf  <- function(X, W, predictkfold){
    rf_grf  <- regression_forest(X, W, tune.parameters = "all")
    p.grf  <- predict(rf_grf, predictkfold)$predictions
    return(p.grf)
}

# storing 
predict.mat  <- matrix(0, nrow = nrow(index), ncol = K)
tauk  <- rep(0, K)
tauk_oracle  <- rep(0, K)
weighttau  <- rep(0, K)
score  <-  list()
score_oracle  <- list()

# for each fold i use other folds for estimation 
for(i in seq(1:K)){
    predict.mat[, i]  <- fun.rf.grf(X = X[c(index[, -i]), ], W = W[index[, -i]], predictkfold = X[c(index[, i]), ])
    # fold-specific treatment effect
    score[[i]]  <- ((W[index[, i]] * Y[index[, i]]) / (predict.mat[, i])) - 
                (((1 - W[index[, i]]) * Y[index[, i]]) / (1 - predict.mat[, i]))
    score_oracle[[i]]  <- ((W[index[, i]] * Y[index[, i]]) / (prob[i])) - 
                (((1 - W[index[, i]]) * Y[index[, i]]) / (1 - prob[i]))

    tauk[i]  <- mean(score[[i]])
    tauk_oracle[i]  <- mean(score_oracle[[i]])
}

# ipw using oracle propensity score and propensity score estimated from grf 
alpha  <-  0.05 # 5 percent level of significance
#ipw.ranger  <- mean(((W * Y) / (p.ranger)) - (((1 - W) * Y) / (1 - p.ranger))) 
ipw.grf  <- weighted.mean(tauk, weights = weighttau)
ipw.oracle  <- weighted.mean(tauk_oracle, weights = weighttau)

sd.ipw  <-  sd(unlist(score))
sd.oracle  <-  sd(unlist(score_oracle))
ll  <- ipw.grf - (sd.ipw / sqrt(length(unlist(score)))) * qnorm(1 - alpha/2)
ul  <- ipw.grf + (sd.ipw / sqrt(length(unlist(score)))) * qnorm(1 - alpha/2)
ll_oracle  <- ipw.oracle - (sd.oracle / sqrt(length(unlist(score_oracle)))) * qnorm(1 - alpha/2)
ul_oracle  <- ipw.oracle + (sd.oracle / sqrt(length(unlist(score_oracle)))) * qnorm(1 - alpha/2)

result.ipw  <- c("IPW estimate" = round(ipw.grf, 3), "se" = round(sd.ipw / (sqrt(length(W))), 3),
                 "lower bound" = round(ll, 3), "upper bound" = round(ul, 3))

result.oracle.ipw  <- c("IPW Oracle estimate" = round(ipw.oracle, 3), "se" = round(sd.oracle / (sqrt(length(W))), 3),
                 "lower bound" = round(ll_oracle, 3), "upper bound" = round(ul_oracle, 3))                 

print(result.ipw)
print(result.oracle.ipw)
```

Note that $e(x)$ is estimated via cross-fitting.
1. The data is divided into $K$-folds. 
2. For each fold $k$, model building is administered using $-k$ folds. 
3. Using Step 2, predictions are generated for units in the $k^{th}$ fold.  

## AIPW and Estimation

Augmented Inverse Probability Weighting (AIPW) provides a robust way to estimate ATE by alleviating the limitation of IPW estimate. 

Following the IPW approach, estimation of ATE is given in equation (6). The other approach to estimate $\tau$ is to think of it from the conditional response approach. Write $\mu_{w}(x) = E[Y_i| \; X_i = x, W_i = w]$. Then: 

$\tau(x) = E[Y_i| \; X_i = x, W_i = 1] - E[Y_i| \; X_i = x, W_i = 0]$


This is the regression outcome approach, where $\tau = E[\mu_{1}(x) - \mu_{0}(x)]$. The consistent estimator can be formed by using: $\hat{\tau}(x) = N^{-1} \sum_{i = 1}^{N} \mu_{1}(X_i) - \mu_{0}(X_i)$. 

AIPW approach combines both IPW approach as well as regression outcome approach to estimate $\tau$. 

$\hat{\tau}_{AIPW} = \frac{1}{N} \sum_{i = 1}^{N} (\mu_{1}(X_i) - \mu_{0}(X_i) + 
\frac{(Y_i - \hat{\mu}_1(X_i)). W_i}{\hat{e}(X_i)} - \frac{(Y_i - \hat{\mu}_0(X_i)). (1-W_i)}{1 - \hat{e}(X_i)})$


ML approach using cross-fitting is used to estimate both $\hat{e}(x)$ and $\hat{\mu}_{w}(x)$. Following the cross-fitting structure, we can formally write the estimate for $\tau$ as: 

$\hat{\tau}_{AIPW} = \lowerbracket{\frac{1}{N} \sum_{i = 1}^{N} (\mu_{1}^{-k(i)}(X_i) - \mu_{0}^{-k(i)}(X_i)}_{consistent \; estimate \; of \; \tau} + 
\frac{(Y_i - \hat{\mu}_1^{-k(i)}(X_i)). W_i}{\hat{e}^{-k(i)}(X_i)} - \frac{(Y_i - \hat{\mu}_0^{-k(i)}(X_i)). (1-W_i)}{1 - \hat{e}^{-k(i)}(X_i)})$

The AIPW approach can be thought of estimating ATE taking the difference across conditional responses. Next, the residuals are adjusted using weights  given by the propensity score. There are two attractive features of AIPW estimate. First, $\hat{\tau}_{AIPW} is consistent as long as $\hat{e}(x)$ or $\hat{\mu}_{w}(x)$ is consistent. This is because $E[(Y_i - \hat{\mu}_{W_i}(X_i)) \approx 0$. Second, $\hat{\tau}_{AIPW}$ is a good approximation to oracle $\hat{\tau}_{AIPW}^{*}$ as long as $\hat{\mu}(.)$ and $\hat{e}(.)$ are reasonably accurate. If one estimate is highly accurate, then it can compensate lack of accuracy on the other estimate. If both $\hat{\mu}(.)$ and $\hat{e}(.)$ are $\sqrt{n}$-consistent^[This means that $\hat{\mu}(.)$ converges to $\hat{\mu}$ at the ], then the following holds. 

$\sqrt{n}(\hat{\tau}_{AIPW} - \hat{\tau}_{AIPW}^{*}) \rightarrow_p 0$.

```{r}

#######################
#
# Augmented IPW (aipw)
#
#######################

#n_features2  <- length(setdiff(names(dat2), "Y"))
# ranger
#funrf_ranger  <- function(dat){
#    rf2  <- ranger(
#        Y ~ ., 
#        data = dat, 
#        mtry = min(ceiling(sqrt(n_features) + 20), n_features),  
#        respect.unordered.factors = "order", 
#        seed = 123, 
#        num.trees = 2000
#    )
#    return(rf2)
#}

# storing
predict.mat2a  <- matrix(0, nrow = nrow(index), ncol = K)
predict.mat2b  <- predict.mat2a
aipwK  <- rep(0, K)
weightaipK  <- rep(nrow(index) / length(index), K)

for(i in seq(1:K)){
    # E(Y | X, W = 1) using cross-fitting
    predict.mat2a[, i]  <- fun.rf.grf(X = cbind(X[c(index[, -i]), ], W[index[, -i]]), W = Y[index[, -i]], 
                                    predictkfold = cbind(X[c(index[, i]), ], 1))
    # E(Y | X, W = 0) using cross-fitting
    predict.mat2b[, i]  <- fun.rf.grf(X = cbind(X[c(index[, -i]), ], W[index[, -i]]), W = Y[index[, -i]], 
                                    predictkfold = cbind(X[c(index[, i]), ], 0))                                
    noise  <-  ((W[index[, i]] * (Y[index[, i]] - predict.mat2a[, i])) / (predict.mat[, i])) 
                - 
                (((1 - W[index[, i]]) * (Y[index[, i]] - predict.mat2b)) / (1 - predict.mat[, i]))
    score[[i]]  <-  predict.mat2a[, i] - predict.mat2b[, i] + noise
    aipwK[i]  <- mean(score[[i]])
}

aipw.grf  <- weighted.mean(aipwK, weights = weightaipK)
sd.aipw  <- sd(unlist(score)) 
ll  <-  aipw.grf - (sd.aipw / sqrt(length(unlist(score)))) * qnorm(1 - alpha/2)
ul  <-  aipw.grf + (sd.aipw / sqrt(length(unlist(score)))) * qnorm(1 - alpha/2)

result.aipw  <- c("AIPW Est." = round(aipw.grf, 3), "se" = round(sd.aipw/(sqrt(length(W))), 3), 
                    "lower bound" = round(ll, 3), "upper bound" = round(ul, 3))

######################
# grf
######################

# Train a causal forest 
tau.forest  <-  causal_forest(X, Y, W)

# Estimate the conditional average treatment effect on the full sample (CATE).
grf_ate  <- average_treatment_effect(tau.forest, target.sample = "all")
grf_att  <- average_treatment_effect(tau.forest, target.sample = "treated")


##  PRINT ALL

#print(paste0("average treatment effect is: ", round(mean(pmax(X[, 1], 0)), 3)))
print(paste("treatment effects according to naive estimator:", round(mean(Y[which(W == 1)]) - mean(Y[which( W == 0)]), 3), sep = " "))
print(paste("treatment effects according to IPW using", K, "fold cross-fittin:", round(ipw.grf, 3), sep = " "))
print(paste("treatment effects according to IPW oracle:", round(ipw.oracle, 3), sep = " "))
print(paste("treatment effects according to AIPW using", K, "fold cross-fitting:", round(aipw.grf, 3), sep = " "))
print(paste("treatment effects according to GRF:", round(grf_ate[[1]], 3), sep = " "))

print(result.ipw)
print(result.aipw)
print(grf_ate)


```

## Assessing Balance
```{r}

##########################################
#
#
# Assessing Balance
#
#
##########################################
XX  <- X[c(index), ]
YY  <- Y[c(index)]
WW  <- W[c(index)]
e.hat  <- c(predict.mat)

# unadjusted
means.treat  <- apply(XX[WW == 1, ], 2, mean)
means.control  <- apply(XX[WW == 0, ], 2, mean)
abs.mean.diff  <- abs(means.treat - means.control)

var.treat  <- apply(XX[WW == 1, ], 2, var)
var.control  <- apply(XX[WW == 0, ], 2, var)
std  <- sqrt(var.treat + var.control)

# adjusted

means.treat.adj  <- apply(XX*WW / e.hat, 2, mean)
means.control.adj  <- apply(XX*(1 - WW) / (1 - e.hat), 2, mean)
abs.mean.diff.adj  <- abs(means.treat.adj - means.control.adj)

var.treat.adj  <- apply(XX * WW / e.hat, 2, var)
var.control.adj  <- apply(XX * (1 - WW) / (1 - e.hat), 2, var)
std.adj  <- sqrt(var.treat.adj + var.control.adj)

# plot unadjusted and adjusted differences
par(oma=c(0,4,0,0))
plot(-2, xaxt="n", yaxt="n", xlab="", ylab="", xlim=c(-.01, 1.01), ylim=c(0, ncol(XX)+1), main="")
axis(side=1, at=c(-1, 0, 1), las=1)
lines(abs.mean.diff / std, seq(1, ncol(XX)), type="p", col="blue", pch=19)
lines(abs.mean.diff.adj / std.adj, seq(1, ncol(XX)), type="p", col="orange", pch=19)
legend("topright", c("Unadjusted", "Adjusted"), col=c("blue", "orange"), pch=19)
abline(v = seq(0, 1, by=.25), lty = 2, col = "grey", lwd=.5)
abline(h = 1:ncol(XX),  lty = 2, col = "grey", lwd=.5)
mtext(paste0("X", seq(1, ncol(XX))), side=2, cex=0.7, at=1:ncol(XX), padj=.4, adj=1, col="black", las=1, line=.3)
abline(v = 0)

hist(e.hat, breaks = 100, freq = FALSE)

```


